Racial and Ethnic Differences in ETI

### 2023-08-29 Kamyron Jordan/Racial and Ethnic Differences in ETI/racial_and_ethnic_differences_in_eti.qmd
cahceing figured out race and eti

```{r data cleaning}
#| include: false
source("/Users/timvigers/GitHub/CF/Kamyron Jordan/Racial and Ethnic Differences in ETI/create_race_eti_analysis_dataset.R")
load("/Volumes/Work/Vigers/CF/Kamyron Jordan/Racial and Ethnic Differences in ETI/Data_Cleaned/analysis_dataset.RData")
```

```{r libraries}
#| include: false
library(Hmisc)
library(tidyverse)
library(gtsummary)
```

@@

```{r table 1}
#| label: tbl-1
#| tbl-cap: Race by year
annual %>%
  select(ReviewYear, race, Hispanicrace) %>%
  tbl_summary(by = ReviewYear, missing_text = "Missing")
```

```{r table 2}
#| label: tbl-2
#| tbl-cap: Race and ethnicity
demo %>%
  select(race, Hispanicrace) %>%
  tbl_summary(by = Hispanicrace) %>%
  modify_header(label = paste0("**", label(annual$Hispanicrace), "**")) %>%
  add_overall()
```

Based on the above table, I decided to combine race and ethnicity with Hispanic "overriding" other racial identities. So if a participant was reported as both Black and Hispanic, they were consider Hispanic for the following analyses. Those with missing data for Hispanic ethnicity were moved to the "Mixed/Other/Unknown Race" category.

```{r combine race and eth}
demo$race <- as.character(demo$race)
demo$race[demo$Hispanicrace == "Yes"] <- "Hispanic or Latino"
demo$race[demo$Hispanicrace == "Unknown"] <- "Hispanic or Latino"
demo$race <- factor(demo$race,
  levels = c(
    "White", "Black or African American", "Hispanic or Latino",
    "American Indian or Alaska Native", "Asian",
    "Native Hawaiian or Other Pacific Islander", "Mixed/Other/Unknown Race"
  )
)
label(demo$race) <- "Race/Ethnicity"
```

@@

```{r}
#| label: tbl-3
#| tbl-cap: Characteristics by race
demo %>%
  select(race) %>%
  tbl_summary()
```

# Figure

```{r}
#| label: fig-planets
#| fig-cap: Planets
```

>>>>

```{r libraries}
#| include: false
library(Hmisc)
library(tidyverse)
library(multidplyr)
library(gtsummary)
library(lmerTest)
library(broom.mixed)
library(emmeans)
library(knitr)
```

```{r data cleaning}
#| include: false
# source("/Users/timvigers/GitHub/CF/Kamyron Jordan/Racial and Ethnic Differences in ETI/create_race_eti_analysis_dataset.R")
load("~/Documents/Work/Vigers/CF/Kamyron Jordan/Racial and Ethnic Differences in ETI/Data_Cleaned/analysis_dataset.RData")
# Make modulators ordered factors
encounter <- encounter %>%
  mutate(across(contains("vx"), ~ factor(.x, ordered = T)))
```

@@

In the tables below, race and ethnicity are considered separately, so some participants may be double counted. For example, a participant identifying as Black and Hispanic would contribute to both categories.

## Race and ethnicity overall

The table below shows the race and ethnicity proportions for the entire cohort included in the CFFPR dataset.

```{r}
demo %>%
  select(race, Hispanicrace) %>%
  tbl_summary(
    by = Hispanicrace,
    statistic = list(all_continuous() ~ "{mean} ({sd})")
  ) %>%
  add_overall() %>%
  bold_labels() %>%
  as_kable_extra()
```

## Race by year

The table below shows the proportions of race and ethnicities included in the dataset in a given year. Most people are seen at least once a year, but this table serves as a sanity check to make sure that groups were generally seen equal amounts each year.

```{r}
annual %>%
  select(ReviewYear, race, Hispanicrace) %>%
  tbl_summary(
    by = ReviewYear, missing_text = "Missing",
    statistic = list(all_continuous() ~ "{mean} ({sd})")
  ) %>%
  bold_labels() %>%
  as_kable_extra()
```

## Race percentages overall

Based on the above tables, I decided to combine race and ethnicity with Hispanic "overriding" other racial identities. So if a participant was reported as both Black and Hispanic, they were consider Hispanic for the following analyses. Those with missing data for Hispanic ethnicity were moved to the "Indigenous/Mixed/Other/Unknown Race" category.

```{r}
demo$race <- as.character(demo$race)
demo$race[demo$Hispanicrace == "Yes"] <- "Hispanic or Latino"
demo$race[demo$Hispanicrace == "Unknown"] <- "Hispanic or Latino"
demo$race <- factor(demo$race,
  levels = c(
    "White", "Black or African American", "Hispanic or Latino", "Asian",
    "American Indian or Alaska Native",
    "Native Hawaiian or Other Pacific Islander", "Mixed/Other/Unknown Race"
  ),
  labels = c(
    "White", "Black or African American", "Hispanic or Latino", "Asian",
    "Indigenous/Mixed/Other/Unknown Race",
    "Indigenous/Mixed/Other/Unknown Race", "Indigenous/Mixed/Other/Unknown Race"
  )
)
encounter$race <- NULL
encounter <- left_join(encounter, demo[, c("eDWID", "race")], by = join_by(eDWID))
label(encounter$race) <- "Race/Ethnicity"
label(demo$race) <- "Race/Ethnicity"
```

```{r}
demo %>%
  select(race) %>%
  tbl_summary(statistic = list(all_continuous() ~ "{mean} ({sd})")) %>%
  bold_labels() %>%
  as_kable_extra()
```

@@

## Characteristics by race

```{r}
demo %>%
  select(race, Gender, WasGenotyped, F508_num) %>%
  tbl_summary(
    by = race, missing_text = "Missing",
    statistic = list(all_continuous() ~ "{mean} ({sd})")
  ) %>%
  add_overall() %>%
  bold_labels() %>%
  as_kable_extra()
```

## Characteristics by year

```{r}
#| cache: true
encounter %>%
  group_by(eDWID, reviewyear) %>%
  summarise(across(c(race, Gender, WasGenotyped, F508_num, contains("insurance")), first),
    `Age at Last Visit` = last(na.omit(encounterage)),
    `Mean ppFEV` = mean(GLI_FEV1_pct_predicted, na.rm = T),
    `Weight at Last Visit` = last(na.omit(weight)),
    `Weight %ile at Last Visit` = last(na.omit(weightpercentile)),
    `BMI at Last Visit` = last(na.omit(bmivalue)),
    `BMI %ile at Last Visit` = last(na.omit(bmipercentile)),
    across(contains("Vx"), ~ max(.x)),
    .groups = "drop"
  ) %>%
  select(
    -eDWID, reviewyear, race, Gender, WasGenotyped, F508_num, contains("insurance"),
    `Age at Last Visit`, `Mean ppFEV`, `Weight at Last Visit`,
    `Weight %ile at Last Visit`, `BMI at Last Visit`,
    `BMI %ile at Last Visit`, contains("Vx")
  ) %>%
  rename(any_of(tidy_labels)) %>%
  tbl_summary(
    by = Year, missing_text = "Missing",
    statistic = list(all_continuous() ~ "{mean} ({sd})")
  ) %>%
  bold_labels() %>%
  as_kable_extra()
```

# Potential COVID effects

## Spaghetti plot

```{r}
#| warning: false
ggplot(encounter, aes(
  y = GLI_FEV1_pct_predicted, x = encounterdate,
  group = eDWID, color = factor(eDWID)
)) +
  geom_line(alpha = 0.02) +
  theme_bw() +
  theme(legend.position = "none")
```

## Visits prior to and after March 11, 2020

The table below is limited to visits 2019 - 2021.

```{r}
encounter %>%
  filter(encounterdate >= "2019-01-01") %>%
  mutate(post_covid = encounterdate > "2020-03-11") %>%
  group_by(eDWID) %>%
  summarise(
    `Number of Visits Pre-COVID` = sum(!post_covid),
    `Number of Visits Post-COVID` = sum(post_covid),
    race = first(race),
    .groups = "drop"
  ) %>%
  rename(any_of(tidy_labels)) %>%
  select(-eDWID) %>%
  tbl_summary(by = `Race/Ethnicity`, missing_text = "Missing") %>%
  bold_labels() %>%
  as_kable_extra()
```

# Lung function over time

## Annualized averages

```{r}
mod_df <- encounter %>%
  group_by(eDWID, reviewyear) %>%
  summarise(
    across(
      c(GLI_FEV1_pct_predicted, GLI_FVC_pct_predicted),
      ~ mean(.x, na.rm = T)
    ),
    race = first(race),
    .groups = "drop"
  )
```

FEV1 and FVC percent predicted (based on GLI equations) were averaged by year for each person.

```{r}
#| warning: false
encounter %>%
  group_by(eDWID, reviewyear) %>%
  summarise(across(c(GLI_FEV1_pct_predicted, GLI_FVC_pct_predicted), ~ mean(.x, na.rm = T)),
    .groups = "drop"
  ) %>%
  pivot_longer(c("GLI_FEV1_pct_predicted", "GLI_FVC_pct_predicted")) %>%
  ggplot(aes(x = reviewyear, y = value, color = factor(eDWID), group = eDWID)) +
  geom_line(alpha = 0.01) +
  theme_bw() +
  theme(legend.position = "none") +
  facet_wrap(~name)
```

```{r}
#| cache: true
# Most basic model of lung function over time by race
mod <- lmer(
  GLI_FEV1_pct_predicted ~ factor(reviewyear) * race + (1 | eDWID),
  data = encounter %>% group_by(eDWID, reviewyear) %>%
    summarise(across(c(GLI_FEV1_pct_predicted, GLI_FVC_pct_predicted), ~ mean(.x, na.rm = T)),
      race = first(race),
      .groups = "drop"
    )
)
res <- tidy(mod, effects = "fixed")
means <- data.frame(emmeans(mod, ~ race + reviewyear,
  lmerTest.limit = 1e6,
  pbkrtest.limit = 1e6
))
kable(means)
```

```{r}
#| fig-width: 10
#| fig-height: 10
ggplot(means, aes(x = reviewyear, y = emmean, color = race)) +
  geom_line() +
  geom_point() +
  xlab("Year") +
  ylab("LS Mean FEV1") +
  geom_errorbar(aes(ymin = emmean - SE, ymax = emmean + SE), width = .2) +
  theme_bw()
```
